Tropical mathematics for the Lambda-calculus 1

### 2023-11-27 CSL24/ArXiv/section8.tex
commit

%In our opinion,
The main goals of this paper are two. Firstly,  to
demonstrate the existence of a conceptual bridge between two different well-studied quantitative approaches to higher-order programs, and to highlight the possibility of transferring results and techniques from one approach to the other. 
Secondly, to suggest that tropical mathematics, a
field which has been largely and successfully applied in computer science, could be used to study quantitative properties of higher-order programs.

While the first goal was here developed in detail, and at different levels of abstraction, for the second goal we only sketched a few interesting directions (best-case analysis, log-probabilities), and we leave their development to a second paper of this series. 
To conclude, while the main ideas of this article only use basic concepts from the toolbox of tropical mathematics, it looks like an exciting direction to look for applications of more advanced tools from tropical algebraic and differential geometry (e.g.~Newton polytopes, tropical varieties, tropical differential equations). 
%We believe that exploring these ideas in more depth could be a fruitful direction; moreover, 
%since both generalized metrics and quantale-modules have been largely studied in computer science, 
%a natural question is if the generalized approach of Section \ref{section6} could  lead to new applications of metric and tropical methods to the $\lambda$-calculus.

>>>>

%In our opinion,
The main goals of this paper are two. Firstly,  to
demonstrate the existence of a conceptual bridge between two well-studied quantitative approaches to higher-order programs, and to highlight the possibility of transferring results and techniques from one approach to the other. 
Secondly, to suggest that tropical mathematics, a
field which has been largely and successfully applied in computer science, could be used for the quantitative analysis of functional programming languages. 
While the first goal was here developed in detail, and at different levels of abstraction, for the second goal we only sketched a few interesting directions, and we leave their development to a second paper of this series.

While the main ideas of this article only use basic concepts from the toolbox of tropical mathematics, an exciting direction is that of looking at potential applications of more advanced tools from tropical algebraic and differential geometry (e.g.~Newton polytopes, tropical varieties, tropical differential equations). Another interesting question is how much of our results on tps and their tropical Taylor expansion can be extended to the abstract setting of generalized metric spaces and continuous functors. 
%We believe that exploring these ideas in more depth could be a fruitful direction; moreover, 
%since both generalized metrics and quantale-modules have been largely studied in computer science, 
%a natural question is if the generalized approach of Section \ref{section6} could  lead to new applications of metric and tropical methods to the $\lambda$-calculus.

### 2023-11-27 CSL24bis/main.tex
commit

\usetikzlibrary{intersections,through}

@@

\title{Tropical Mathematics and the Lambda-calculus I}
\subtitle{Metric and Differential Analysis of Effectful Programs
} %TODO Please add

\titlerunning{Tropical mathematics for the Lambda-calculus I} %TODO optional, please use if title is longer than one line

\author{Davide {Barbarossa}}{Dipartimento di Informatica - Scienza e Ingegneria, Universit\`a di Bologna, %[optional: Address], 
Italy% \and My second affiliation, Country 
\and \url{https://lipn.univ-paris13.fr/\~barbarossa/index.html} }{davide.barbarossa@unibo.it}{https://orcid.org/0000-0003-4608-8282}{%(Optional)Funding each author
}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{Paolo {Pistone}}{Dipartimento di Informatica - Scienza e Ingegneria, Universit\`a di Bologna, %[optional: Address], 
Italy% \and LIP, Universit\'e de Lyon, France 
\and \url{http://logica.uniroma3.it/pistone/
} }{paolo.pistone2@unibo.it}{https://orcid.org/0000-0003-4250-9051}{%(Optional)Funding each author
}

@@

\keywords{Relational semantics, Differential lambda-calculus, Tropical semiring,  Program metrics, Lawvere quantale} %TODO mandatory; please add comma-separated list of keywords

@@

\relatedversion{{\color{red}ADD URL HERE.}} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

@@

\funding{This work has been supported by the European Research Council through the project ERC CoG 818616 DIAPASoN.}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
We study the interpretation of the lambda-calculus in a framework based on tropical mathematics, and we show that it provides a unifying framework for two well-developed quantitative approaches to program semantics: on the one hand program metrics, based on the analysis of program sensitivity via Lipschitz conditions, on the other hand resource analysis, based on linear logic and higher-order program differentiation. 
To do that we focus on the semantics arising from the relational model weighted over the tropical semiring, and we discuss its application to the study of “best case” program behavior for languages with probabilistic and non-deterministic effects. Finally, we show that a general foundation for this approach is provided by an abstract correspondence between tropical algebra and Lawvere’s theory of generalized metric spaces.
%A typical result of such line of research is that, thanks to the tropical operations, a simply typed term is interpreted as a locally Lipschitz map, which is in turn decomposed as an inf of more and more sensitive Lipschitz ones, corresponding to the interpretations of the approximants given by the lambda-calculus Taylor expansion.
%We finally describe some promising lines of research, showing in particular how such tropical model expresses natural optimisation problems of probabilistic calculi.
\end{abstract}

@@

\section{A Bridge between Metric and Differential Aspects}\label{section5bis}
\input{section5bis}

\section{Tropical Polynomials and Power Series}\label{section2}
\input{section2bis}

\section{Tropical Semantics and First Order Effectful Programs}\label{section22}
\input{section2bisbis}

\section{Tropical Semantics of Higher-Order Programs}\label{section3}
\input{section3fancy}

%%
%%\section{Topical linear/non-linear algebra}\label{section3}
%%\input{section3C}
%%
%%\section{Tropical weighted relational semantics for high-order languages}
%%
%%The category $\LREL$ satisfies many nice properties which make it able to model all the variants of the $\lambda$-calculi to which we are interested in here. Although the majority of such properties are shared with the general class of models of $\lambda$-calculi called \emph{weighted relational models}, thoroughly studied in \cite{Manzo2013}, we mention them for our particular case in order to highlight how these look like in the tropical setting, but also to make the non-expert in such topics reader able to follow.
%%The real interest in our particular model will come in the next \autoref{sec:tls}.
%%
%%
%%\input{section3A}
%%
%%\input{section2}
%
%\paragraph*{Allowed duplication/erasure: the probabilistic $\lambda$-calculus $\STLC_\oplus$}\label{sec:STLCO}
%
%\input{section3B}

\section{On Tropical Power Series}\label{sec:tls}
\input{section4}

\section{Lipschitz Meets Taylor}\label{sec:TayLip}
\input{sectionTaylip}

\section{Generalized Metric Spaces and $\Lawv$-Modules}\label{sec:GMS}
\input{sectionLaw}

@@

\section{Conclusion and Future Work}\label{section8} \input{section8}

@@

\bibliography{tropical}

\appendix

\section{Proofs from \autoref{section2}: Theorem \ref{theorem:fepsilon}}
\input{appSection3}

\section{Proofs from \autoref{section3}: $(\LREL, !_{n})$ is a model of $\BSTLC$}
\input{appSection5}

\section{Proofs from \autoref{sec:tls}: the Analysis of Tropical Power Series}\label{appsec:tls}
\input{app_section4}

\section{Proofs from \autoref{sec:TayLip}: Lipschitz meets Taylor}
\input{appSection7}

\section{Proofs of \autoref{sec:GMS}: $\Lawv$-modules and Generalized Metric Spaces}
\input{appSection8}

>>>>

@@

\title{Tropical mathematics for the Lambda-calculus 1}
\subtitle{The potential of tropical wheighted relational semantics%for probabilities and Taylor expansion
} %TODO Please add

\titlerunning{Tropical mathematics for the Lambda-calculus 1} %TODO optional, please use if title is longer than one line

\author{Davide {Barbarossa}}{Dipartimento di Informatica, Universit\`a di Bologna, %[optional: Address], 
Italy% \and My second affiliation, Country 
\and \url{https://lipn.univ-paris13.fr/\~barbarossa/index.html} }{davide.barbarossa@unibo.it}{https://orcid.org/0000-0003-4608-8282}{%(Optional)Funding each author
}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{Paolo {Pistone}}{Dipartimento di Informatica, Universit\`a di Bologna, %[optional: Address], 
Italy% \and LIP, Universit\'e de Lyon, France 
\and \url{%WebPage
} }{paolo.pistone2@unibo.it}{%Orchid
}{%(Optional)Funding each author
}

@@

\keywords{Differential lambda-calculus, Tropical semiring, Relational semantics, Lawvere quantale, Program metrics} %TODO mandatory; please add comma-separated list of keywords

@@

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

@@

\funding{This work has been supported by the DIAPASoN project}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
We study the interpretation of the lambda-calculus in a framework based on tropical mathematics, by considering the relational semantics weighted on the tropical semiring. The aim is to provide a framework where both program metrics, based on the analysis of program sensitivity via Lipschitz conditions, and resource analysis, based on higher-order program differentiation, coexist. In order to do that, we consider a graded-types lambda-calculus and the differential lambda-calculus, both languages ensuring bounded duplications.
A typical result of such line of research is that, thanks to the tropical operations, a simply typed term is interpreted as a locally Lipschitz map, which is in turn decomposed as an inf of more and more sensitive Lipschitz ones, corresponding to the interpretations of the approximants given by the lambda-calculus Taylor expansion.
We finally describe some promising lines of research, showing in particular how such tropical model expresses natural optimisation problems of probabilistic calculi.
\end{abstract}

@@

\section{Preliminaries on Tropical (Linear and Non-Linear) Algebra}
\input{section2bis}

\section{Tropical Semantics of Effectful Programs: Two Examples}
\input{section2bisbis}

\section{Tropical Semantics of Higher-Order Programs}\label{section3}

The category $\LREL$ satisfies many nice properties which make it able to model all the variants of the $\lambda$-calculi to which we are interested in here. Although the majority of such properties are shared with the general class of models of $\lambda$-calculi called \emph{weighted relational models}, thoroughly studied in \cite{Manzo2013}, we mention them for our particular case in order to highlight how these look like in the tropical setting, but also to make the non-expert in such topics reader able to follow.
The real interest in our particular model will come in the next \autoref{sec:tls}.

\paragraph*{Prohibited duplication/erasure: the \emph{linear} $\STLC$}\label{sec:3A}
\input{section3A}

\paragraph*{Allowed duplication/erasure: the $\STLC$ and a toy probabilistic $\lambda$-calculus $\STLC_\oplus$}\label{sec:STLC+STLCO}

\input{section3B}

\input{section2}

\section{Tropical Laurent Series}\label{sec:tls}
\input{section4}

@@

\section{Conclusion}\label{section8} \input{section8}

@@

%\bibliography{tropical}

%\appendix

%\section{\autoref{section3}}\label{app-sec:section3}
